---
title: "Red Wine Data Exploration"
author: William Dew
output: html_document
---

========================================================

```{r echo=FALSE, message=FALSE, warning=FALSE, packages}
# Load all of the packages that you end up using in your analysis in this code
# chunk.

# Notice that the parameter "echo" was set to FALSE for this code chunk. This
# prevents the code from displaying in the knitted HTML output. You should set
# echo=FALSE for all code chunks in your file, unless it makes sense for your
# report to show the code that generated a particular plot.

# The other parameters for "message" and "warning" should also be set to FALSE
# for other code chunks once you have verified that each plot comes out as you
# want it to. This will clean up the flow of your report.

library(ggplot2)
library(gridExtra)
library(GGally)
library("Hmisc")
library(corrplot)
library(Simpsons)
```

```{r echo=FALSE}
# Load the Data
wine <- read.csv("C:/Users/wdew/Desktop/jupyter/P4/wineQualityReds.csv", header=TRUE)
wine_q <- wine

#Transforming Quality from an Integer to a Factor
wine$quality <- factor(wine$quality, ordered = T)
```
## Quest of the Project

I will be analyzing data from different bottle of red wine.  I will see if there
is any determining factors that will cause the quality of the wine to be rated 
higher.  First I will explore the variables on there own.  Then I will see if 
there is any corrilation between the variables and the quality of the red wine. 
I will also throw in other factors into the corrilation to see if it effects the
wine qulity.

# Exploring the Data

```{r echo=FALSE}

names(wine)

```

```{r echo=FALSE}

str(wine)

```

```{r echo=FALSE}

summary(wine)

```

### Data Structure
The data set has 1599 rows and 13 columns.  The first column is a unique number 
for each row so it can be discared.  Most of the variables are numerical and 
represent different chemical traits of the red wine.  Quality is a catagorial 
variable that denotes good or bad range of the red wine.  Most wine is average 
with few really good or really bad wines.  This may be a lack of data or it may 
be that most wine is average.

# Univariate Plots Section

### Variable descriptions

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(data = wine, aes(x = quality)) +
  geom_bar(width = 1, color = 'black',fill = I('green'))

```

quality is a normal distribution but a very high peak at 5 and 6.


```{r echo=FALSE}

grid.arrange(ggplot(wine, aes( x = 1, y = fixed.acidity ) ) + 
               geom_boxplot(alpha = 0.2, color = 'green' ),
ggplot(data = wine, aes(x = fixed.acidity)) +
  geom_histogram(binwidth = 1, color = 'black',fill = I('green')), ncol=2)

```

fixed.acidity has a positive skew with a median of 7.5.  The outlires pull 
the mean to 8.32.

```{r echo=FALSE}

grid.arrange(ggplot(wine, aes( x = 1, y = volatile.acidity ) ) + 
               geom_boxplot(alpha = 0.2, color = 'green' ),
ggplot(data = wine, aes(x = volatile.acidity)) +
  geom_histogram(binwidth = 0.05, color = 'black',fill = I('green')), ncol=2)

```

volitile.acidity seems to have a normal distribution but is positivly skewed 
because of some outlires.

```{r echo=FALSE}

grid.arrange(ggplot(wine, aes( x = 1, y = citric.acid ) ) + 
               geom_boxplot(alpha = 0.2, color = 'green' ),
ggplot(data = wine, aes(x = citric.acid)) +
  geom_histogram(binwidth = 0.05, color = 'black',fill = I('green')), ncol=2)

```

citric.acid looks strange.  A lot of bottle didn't have any citric acid and 
then it seemed to be a normal distribution.  Was the data not complete or did a 
lot of sample not have citric acid?

```{r echo=FALSE}

grid.arrange(ggplot(wine, aes( x = 1, y = residual.sugar ) ) + 
               geom_boxplot(alpha = 0.2, color = 'green' ),
ggplot(data = wine, aes(x = residual.sugar)) +
  geom_histogram(binwidth = 1, color = 'black',fill = I('green')), ncol=2)

```

```{r echo=FALSE}

grid.arrange(ggplot(wine, aes( x = 1, y = chlorides ) ) + 
               geom_boxplot(alpha = 0.2, color = 'green' ),
ggplot(data = wine, aes(x = chlorides)) +
  geom_histogram(binwidth = 0.01, color = 'black',fill = I('green')), ncol=2)

```

chloride has a lot of outliers making it positively skewed but has one tall 
peak at 0.05.

```{r echo=FALSE}

grid.arrange(ggplot(wine, aes( x = 1, y = free.sulfur.dioxide ) ) + 
               geom_boxplot(alpha = 0.2, color = 'green' ),
ggplot(data = wine, aes(x = free.sulfur.dioxide )) +
  geom_histogram(binwidth = 1, color = 'black',fill = I('green')), ncol=2)

```

free.sulfur.dioxide has one tall peek then goes positively skewed from 
outliers

```{r echo=FALSE}

grid.arrange(ggplot(wine, aes( x = 1, y = total.sulfur.dioxide ) ) + 
               geom_boxplot(alpha = 0.2, color = 'green' ),
ggplot(data = wine, aes(x = total.sulfur.dioxide )) +
  geom_histogram(binwidth = 5, color = 'black',fill = I('green')), ncol=2)

```

total.sulfur.dioxide has the same tall peek then goes positively skewed like 
the patterns we have been seeing.

```{r echo=FALSE}

grid.arrange(ggplot(wine, aes( x = 1, y = density ) ) + 
               geom_boxplot(alpha = 0.2, color = 'green' ),
ggplot(data = wine, aes(x = density)) +
  geom_histogram(binwidth = 0.001, color = 'black',fill = I('green')), ncol=2)

```

density has a normal distribution.

```{r echo=FALSE}

grid.arrange(ggplot(wine, aes( x = 1, y = pH ) ) + 
               geom_boxplot(alpha = 0.2, color = 'green' ),
ggplot(data = wine, aes(x = pH)) +
  geom_histogram(binwidth = 0.1, color = 'black',fill = I('green')), ncol=2)

```

ph has a normal distribution.

```{r echo=FALSE}

grid.arrange(ggplot(wine, aes( x = 1, y = sulphates ) ) + 
               geom_boxplot(alpha = 0.2, color = 'green' ),
ggplot(data = wine, aes(x = sulphates)) +
  geom_histogram(binwidth = 0.1, color = 'black',fill = I('green')), ncol=2)

```

sulphates has the same positively skewed tail of outliers that a lot of the 
other variables have.

```{r echo=FALSE}

grid.arrange(ggplot(wine, aes( x = 1, y = alcohol ) ) + 
               geom_boxplot(alpha = 0.2, color = 'green' ),
ggplot(data = wine, aes(x = alcohol)) +
  geom_histogram(binwidth = 0.1, color = 'black',fill = I('green')), ncol=2)

```

alchohol has a postivively skewed curve but doesn't have the tail of 
outliers.

# Univariate Analysis

### Feature of Interest

The effect different variables have on quality. 

### Hypothosis

I don't drink wine so I can only guess at what makes wine taste better and have 
a better quality.  I suspect that more alcohol will increase quality.  Residual 
Sugar will have a positive impact of quality because sugar helps taste.

# Bivariate Plots Section

The first step I will take is to make a corrilation table to see where I will 
start my investigation.  Finding high positive or negative corrilations will 
point me in the right dirction.

```{r echo=FALSE}

res2 <- rcorr(as.matrix(wine))
res <- cor(wine_q)

corrplot(res, type = "upper", order = "hclust", 
         tl.col = "black", tl.srt = 45)

```

From the graph above we learn some things.
1. quality is most corillated with alcohol and volatile.acid
2. quality is corrilated with citric.acid and sulphates.
3. fixed.acidity has a very strong corrilation with pH.
4. pH and volatile.acid have a positive corrilation which is weird 
becasue acid has a negative corrilation with pH.

# Bivariate Analysis

I will now create some boxplots to see if I have missed anything from the correlation table.


```{r echo=FALSE}

ggplot(data = wine, aes(x = quality, y = alcohol)) +
  geom_jitter( alpha = .3) +
  geom_boxplot(alpha = .5,color = 'green') +
  stat_summary(fun.y = "mean", 
               geom = "point", 
               color = "red", 
               shape = 8, 
               size = 4)

```


1. Alcohol and Quality has the best positive correlation at 0.4761. As you can see from 
the graph above as alcohol content goes up the quality of the wine goes up.

```{r echo=FALSE}

ggplot(data = wine, aes(x = quality, y = volatile.acidity)) +
  geom_jitter( alpha = .3) +
  geom_boxplot(alpha = .5,color = 'green') +
  stat_summary(fun.y = "mean", 
               geom = "point", 
               color = "red", 
               shape = 8, 
               size = 4)

```

2. volatile.acidity and Quality have the best negative correlation at -0.3905.  As 
volatile.acidity goes down the quality of wine goes up.

```{r echo=FALSE}

ggplot(data = wine, aes(x = quality, y = sulphates)) +
  geom_jitter( alpha = .3) +
  geom_boxplot(alpha = .5,color = 'green') +
  stat_summary(fun.y = "mean", 
               geom = "point", 
               color = "red", 
               shape = 8, 
               size = 4)

```

3.  sulphates has a positive corrilation with quality at 0.2413 as seen in the graph 
above.

```{r echo=FALSE}

ggplot(data = wine, aes(x = quality, y = citric.acid)) +
  geom_jitter( alpha = .3) +
  geom_boxplot(alpha = .5,color = 'green') +
  stat_summary(fun.y = "mean", 
               geom = "point", 
               color = "red", 
               shape = 8, 
               size = 4)

```

4. citric.acid has a positive corrilation with quality at 0.2263 as shown in the graph 
above.

I wan to investigate individual acid on pH because of the negative correlation
between volatile.acid and pH.

```{r echo=FALSE}

ggplot(data = wine, aes(x = fixed.acidity, y = pH)) +
  geom_point(alpha = 0.3) + 
  geom_smooth(method="lm")

```

```{r echo=FALSE}

ggplot(data = wine, aes(x = citric.acid, y = pH)) +
  geom_point(alpha = 0.3) + 
  geom_smooth(method="lm")

```

```{r echo=FALSE}

ggplot(data = wine, aes(x = volatile.acidity, y = pH)) +
  geom_point(alpha = 0.3) + 
  geom_smooth(method="lm")

```

5. pH has a negative correlation with fixed.acidity and citric.acid.  This 
makes sence since low pH numbers means more acidic.  The more citric.acid in the
wine the lower the pH. But volitile.acidity and pH have a positive correlation.  
This is really weird because the more acid you have should decrease the you pH.  
I will need to look into this using the simpons paradox and see if there is 
another variable that is changing the coefficient.

```{r echo=FALSE, message=FALSE, warning=FALSE}

simpsons <- Simpsons(volatile.acidity, pH, data=wine)
plot(simpsons)

```

I clustered the data into 3 segments and calculated the regression coefficient. I see that there is indeed a sign reversal. This is due to a lurking variable which changes the overall coefficient.

# Multivariate Plots Section

Alcohol has the biggest correlation to quality of wine.  We will not look to see
if any other variables contribute to quality of wine if alcohol stays constant.

```{r echo=FALSE}

ggplot(data = wine,
       aes(y = volatile.acidity, x = alcohol,
           color = quality)) +
  geom_point(alpha = 0.8, size = 1) +
  geom_smooth(method = "lm", se = FALSE,size=1)  +
  scale_color_brewer(type='seq',
                   guide=guide_legend(title='Quality'))

```

We can see that less volatile.acidity improves quality when alcohol is constant.

```{r echo=FALSE}

ggplot(data = wine,
       aes(y = citric.acid, x = alcohol,
           color = quality)) +
  geom_point(alpha = 0.8, size = 1) +
  geom_smooth(method = "lm", se = FALSE,size=1)  +
  scale_color_brewer(type='seq',
                   guide=guide_legend(title='Quality'))

```

In the graph above we see that higher quality of wine comes from higher 
citric.acid content when alcohol is constant.

```{r echo=FALSE}

ggplot(data = wine,
       aes(y = sulphates, x = alcohol,
           color = quality)) +
  geom_point(alpha = 0.8, size = 1) +
  geom_smooth(method = "lm", se = FALSE,size=1)  +
  scale_color_brewer(type='seq',
                   guide=guide_legend(title='Quality'))

```

When sulphates and alcohol are at a higher concentration the quality of wine is 
better.

# Multivariate Analysis

1. High Alcohol and Sulaphate content seems to produce better wines.
2. Citric Acid, even though weakly correlated plays a part in improving the wine
quality.

------

# Final Plots and Summary

> **Tip**: You've done a lot of exploration and have built up an understanding
of the structure of and relationships between the variables in your dataset.
Here, you will select three plots from all of your previous exploration to
present here as a summary of some of your most interesting findings. Make sure
that you have refined your selected plots for good titling, axis labels (with
                                                                         units), and good aesthetic choices (e.g. color, transparency). After each plot,
make sure you justify why you chose each plot by describing what it shows.

### Plot One
```{r echo=FALSE}

ggplot(data = wine, aes(x = quality, y = alcohol)) +
  geom_jitter( alpha = .3) +
  geom_boxplot(alpha = .5,color = 'green') +
  stat_summary(fun.y = "mean", 
               geom = "point", 
               color = "red", 
               shape = 8, 
               size = 4) + 
  ggtitle("Quality and Alcohol Boxplot correlation")

```

### Description One

This plot shows us the positive corrilation between alcohol and quality.  The 
higher the alcohol content the higher the quality of wine.  You can also see 
from this graph that the distribution normal with a high median of alcohol 
implies great wine.

### Plot Two
```{r echo=FALSE}

ggplot(data = wine, aes(x = volatile.acidity, y = pH)) +
  geom_point(alpha = 0.3) + 
  geom_smooth(method="lm") + 
  ggtitle("Volatile Acidity and pH Correlation")

```

### Description Two

This plot shows us the weird positive corrilation between pH and 
volatile.acidity.  This didn't make sense because we know that acids have a 
negative correlation with pH.  The more acid the lower the pH.  In order to 
figure this out we looke towards the Simpson paradox and determined there was an
underlying variable that was causing this positive corrilation between pH and 
volatile.acidity.

### Plot Three
```{r echo=FALSE}

ggplot(data = wine,
       aes(y = sulphates, x = alcohol,
           color = quality)) +
  geom_point(alpha = 0.8, size = 1) +
  geom_smooth(method = "lm", se = FALSE,size=1)  +
  scale_color_brewer(type='seq',
                   guide=guide_legend(title='Quality')) +
  ggtitle("Sulphates and Alcohol Correlation by Quality Level")

```

### Description Three

From this graph we can see higher quality of wines contain high values of 
alcohol and sulphates. This implys that higher concentrations of alcohol and 
sulphates increase wine quality.  There is a downward slope, even if is very 
slight that might indicate that higher alcohol is a little better then higher 
sulphates.

------
  
# Reflection
  
So when I first started to look at the data and plot out the variables I noticed
that most of the data was average quality of wine with very few bad and great 
quality of wines.  This helped in showing a normal distribution but it was not a
great representation for trying to determine what variables effected quality of 
wine.  This caused some trouble with some of the variables like citric.acid 
which had a lot of zeros in the data.  I couldn't tell if the data was
incomplete or if there were a lot of wine with no citric.acid in them.

When I started to look at the correlation between the variables a very weird 
result came up.  It seemed that pH and volitile acid had a positve correlation.  
This meant that as more acid was added pH would become more basic.  
This is obviously wrong.  I need to look into this so I started reading about 
this phenomenom and discovered the Simpson paradox.  I was able to apply this 
principle and determine that there were other variables that were causing the 
positive correlation between volatile acid and pH.

Something that would be fun to look into would be how wine determined to be of 
quality by a taster.  You could have a little survey with a couple of reponses 
and see if a certain sense played a factor or if a taster was bias towards 
certain types of wine.
